---
title: "Varpart"
author: "Marion Boisseaux"
date: "2023-05-16"
output: html_document
---
Bray–Curtis distances were used for microbial composition in this analysis.

```{r}
load("./resources/Metabarlist_natura_clean_ITS2_traits_alpha.Rdata")
leaf <- subset_metabarlist(natura_clean, table = "samples",
                          indices = natura_clean$samples$organ == "leaf")
```


### leaf ITS - compi

```{r}
# microbial data, hellinger transformed
microbio.hel <- leaf$reads %>% decostand(., "hell")
microbio.hel.bray <- vegdist(microbio.hel, method = "bray")

#trait data
names(leaf$samples)

trait_st <- leaf$samples %>% dplyr::select(RootShoot, RootLength_Total, SLA, LA_rwc, Height, LSWC, fvfm, StomatalDensity, SRL, RootDiameter, LT, gmin, MEM1, MEM2, MEM3, MEM4) %>% scale(center = TRUE, scale = TRUE) %>% as.data.frame()# Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.
trait_st$sample_id <- rownames(trait_st)
trait_st <- left_join(trait_st, leaf$samples %>% dplyr::select(sample_id, name))


# Calculate variance partitioning, 

#use partial RDA to test significance of fractions of interest
m_complet <- dbrda(microbio.hel.bray ~SLA+ LA_rwc+ Height+ LSWC+ fvfm+ StomatalDensity+ LT+ gmin+ MEM1+ MEM2+ MEM3+ MEM4+ name, data = trait_st) 
RsquareAdj(m_complet) #3%

#extract2 PCA axes
PCA12 <- scores(m_complet, display= "sites", choices = 1:2)
m_complet_PCA12 <- dbrda(microbio.hel.bray ~ PCA12) 
RsquareAdj(m_complet_PCA12) 

#https://www.davidzeleny.net/anadat-r/doku.php/en:expl_var_examples
# Two first tb-PCA axes, if used as explanatory in the db-RDA on the same dataset, explain 11% of the variance. Note that this is the same number we would get by simply checking the amount of variance represented by the first two ordination axes in db-RDA in the summary above: eigenvalue of PCA1 = 0.09197, eigenvalue of PCA1 = 0.06075, total variance (inertia) = 0.70476, recalculated into percentage: (0.09197+0.06075)/0.70476 = 21.7%. In fact, we don't need to really to the whole procedure (extract tb-PCA axes and use them as explanatory in tb-RDA), we can simply check the variance of n axes in unconstrained ordination (n = number of explanatory variables) and compare it with real variance explained by environmental variables.
# 
# The comparison here is: 8.9% explained by real variables (pH and SOILDPT) vs 21.7% which would be explained by two best, not correlated variables (if we had them). We see that measured variables explain something over 40% of variation they could (8.9/21.7 = 0.41), which is not bad. Remember important difference: PCA axes are (from definition) not correlated, while our real variables often will be (as in this case: cor.test (~ pH + SOILDPT, data = env) shows that Pearson's correlation coefficient between pH and SOILDPT is r = 0.273, and this correlation is significant at P = 0.0069).

m_null <- dbrda(microbio.hel.bray ~1, data = trait_st) 
RsquareAdj(m_null)

m_leaf <- dbrda(microbio.hel.bray ~LSWC + LT + gmin + SLA, La_rwc+ StomatalDensity + fvfm, data= trait_st)
RsquareAdj(m_leaf) 

m_plant <- dbrda(microbio.hel.bray ~ Height , data = trait_st)
RsquareAdj(m_plant) 

m_loc <- dbrda(microbio.hel.bray ~ MEM1 + MEM2 + MEM3 +MEM4, data = trait_st)
RsquareAdj(m_loc) 

m_ID <- dbrda(microbio.hel.bray ~ name, data= trait_st)
RsquareAdj(m_ID)

anova(m_complet) #0.001 **
anova(m_leaf) #.
anova(m_plant) #.
anova(m_loc) #0.006 **
anova(m_ID) #0.005**
anova(m_null) #0

#roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, trait_st)[,-1]
# Test fraction [b] using partial RDA:
leaftraits <- model.matrix(~LSWC + LT + gmin + SLA + LA_rwc + StomatalDensity + fvfm, trait_st)[,-1]
# Test fraction [c] using partial RDA:
location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, trait_st)[,-1]
# Test fraction [d] using partial RDA:
identity <- model.matrix(~name, trait_st)[,-1]
# Test fraction [e] using partial RDA:
planttraits <- model.matrix(~Height, trait_st)[,-1]

vp <- varpart(microbio.hel.bray,  ~leaftraits, ~identity, ~location, ~planttraits, data = trait_st)
vp
plot(vp)
```

### leaf ITS - diversite

```{r}
#trait data
trait_st <- leaf_ITS$samples %>% dplyr::select(RootLength_Total , SRL ,RootDiameter , NbLeaves , LSWC , StomatalDensity  , LT , gmin , TLP) %>% decostand(., "standardize") # Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.

#location data


location <- leaf_ITS$samples %>% dplyr::select(MEM1,MEM2, MEM3, MEM4) #pcnm

# Calculate variance partitioning
leafdf2 <- df2 %>%filter(organ == "leaf")

vp2 <-varpart(Y = leafdf2$Shannon ,  ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm, # leaf traits
               #~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = leafdf2)
vp2
showvarparts(3, bg=2:4)
plot(vp2, cex = 2, bg=2:4, Xnames = c("root traits", "leaf traits","species Id"), title(main = "Foliar fungal endophytes"))


#use RDA to test significance of fractions of interest

# Test fraction [a] using partial RDA:
roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, leafdf2)[,-1]
# Test fraction [b] using partial RDA:
leaftraits <- model.matrix(~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm, leafdf2)[,-1]
# Test fraction [c] using partial RDA:
location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, leafdf2)[,-1]
# Test fraction [d] using partial RDA:
identity <- model.matrix(~name, leafdf2)[,-1]

aFrac <- rda(leafdf2$Shannon, roottraits)
bFrac <- rda(leafdf2$Shannon, leaftraits)
cFrac <- rda(leafdf2$Shannon, location)
dFrac <- rda(leafdf2$Shannon, identity)

anova(aFrac, step=200, perm.max=200) #0.007 ** #sans location 0.005 **
anova(bFrac, step=200, perm.max=200) #0.001 *** #sans location 0.002 **
anova(cFrac, step=200, perm.max=200) #0.863
anova(dFrac, step=200, perm.max=200) #0.001 *** #sans location 0.001 **

# RsquareAdj gives the same result as component [a] of varpart
a <- unlist(RsquareAdj(aFrac)[2])
b <- unlist(RsquareAdj(bFrac)[2])
c <- unlist(RsquareAdj(cFrac)[2])
d <- unlist(RsquareAdj(dFrac)[2])

total <- a+b+d

1-total

rda(leafdf2$Shannon  ~SRL + RootDiameter + RootLength_Total + LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm + MEM1 + MEM2 + MEM3 +MEM4+name, #species' ID
               data = leafdf2, perm = 999 )
# 
# Call: rda(formula = leafdf2$Shannon ~ SRL + RootDiameter +
# RootLength_Total + LSWC + LT + gmin + NbLeaves + LA_Total +
# StomatalDensity + fvfm + MEM1 + MEM2 + MEM3 + MEM4 + name, data =
# leafdf2, perm = 999)
# 
#               Inertia Proportion Rank
# Total          0.7559     1.0000     
# Constrained    0.4388     0.5805    1
# Unconstrained  0.3171     0.4195    1
# Inertia is variance 
# 
# Eigenvalues for constrained axes:
#   RDA1 
# 0.4388 
# 
# Eigenvalues for unconstrained axes:
#    PC1 
# 0.3171 

```






Not sure this method is correct... 

```{r}
#null model
model_null <- nlme::lme(Richness~ 1,  random=~1|name, data = df2, na.action = na.omit, method = "ML")

#root traits
model_root_trait <- nlme::lme(Richness ~ SRL + RootDiameter + RootLength_Total,  random=~1|name, data = df2, na.action = na.omit, method = "ML")

#leaf traits
model_leaf_trait <- nlme::lme(Richness  ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm,  random=~1|name, data = df2, na.action = na.omit, method = "ML")

#location
model_pcnm <-  nlme::lme(Richness ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = df2, na.action = na.omit, method = "ML")

# Extract the variance components for each model
Var_components_null <- lme4::VarCorr(model_null)
Var_components_root <- lme4::VarCorr(model_root_trait)
Var_components_leaf <- lme4::VarCorr(model_leaf_trait)
Var_components_pcnm <- lme4::VarCorr(model_pcnm)

# Extract the species (random) effect variances and the residual variance for each model
sp_null <- as.numeric(Var_components_null[1,1])
sp_root <- as.numeric(Var_components_root[1,1])
sp_leaf <- as.numeric(Var_components_leaf[1,1])
sp_pcnm <- as.numeric(Var_components_pcnm[1,1])

res_null <- as.numeric(Var_components_null[2,1])
res_root <- as.numeric(Var_components_root[2,1])
res_leaf <- as.numeric(Var_components_leaf[2,1])
res_pcnm <- as.numeric(Var_components_pcnm[2,1])

# Obtain the total variance of the null model
v_0 <- sp_null + res_null

# Calculate the variance component linked to the root traits
var_species_root <- round(100 * sp_root / v_0)
var_res_root <- round(100 * res_root / v_0)
var_trait_root <- round(100 - (var_species_root + var_res_root)) #only linked to the traits
  
# Calculate the variance component linked to the root traits
var_species_leaf <- round(100 * sp_leaf / v_0)
var_res_leaf <- round(100 * res_leaf / v_0)
var_trait_leaf <- round(100 - (var_species_leaf + var_res_leaf)) #only linked to the traits

# Calculate the variance component linked to the location
var_species_loc <- round(100 * sp_pcnm / v_0)
var_res_loc <- round(100 * res_pcnm / v_0)
var_trait_loc <- round(100 - (var_species_loc + var_res_loc)) #only linked to the traits

#calculate the variance linked to the species identity
var_species_ID <- round(100 * sp_null / v_0)

#calculate residual variance
var_res_total <- round(100- (var_trait_leaf + var_trait_root + var_trait_loc+ var_species_ID))

#organize for plottibng
  Organ <- c("Leaf")
  Diversity <- c("Richness")
  Levels <- c( "Residual", "Species", "Location", "Leaf traits", "Root traits")
  Variances <- c( var_res_total,var_species_ID, var_trait_loc, var_trait_leaf,  var_trait_root)
vars.richness.leaf <-   data.frame(Organ, Diversity, Levels, Variances)

#plot the results
vars.richness.leaf$Levels <- as.factor(vars.richness.leaf$Levels)
variance_plot<- vars.richness.leaf %>%  
  mutate(Levels = factor(Levels, levels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Diversity)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#db7c26",  "#a7c957", "#bc4749", "#62929e", "#BBBBBB" ),
                    breaks=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"),
                    labels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)+
  ggtitle("Fungi Variance partitioning")
variance_plot
```

```{r}

#function for variance partitioning
Var_par <- function(Diversity, Mydata, Organ){

  #Rename the trait column
  colnames(Mydata)[which(colnames(Mydata) == Diversity)] <- "Diversity"  

  #null model
model_null <- nlme::lme(Diversity~ 1,  random=~1|name, data = Mydata, na.action = na.omit, method = "ML")

#root traits
model_root_trait <- nlme::lme(Diversity ~ SRL + RootDiameter + RootLength_Total,  random=~1|name, data = Mydata, na.action = na.omit, method = "ML")

#leaf traits
model_leaf_trait <- nlme::lme(Diversity  ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm,  random=~1|name, data = Mydata, na.action = na.omit, method = "ML")

#location
model_pcnm <-  nlme::lme(Diversity ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = Mydata, na.action = na.omit, method = "ML")

# Extract the variance components for each model
Var_components_null <- lme4::VarCorr(model_null)
Var_components_root <- lme4::VarCorr(model_root_trait)
Var_components_leaf <- lme4::VarCorr(model_leaf_trait)
Var_components_pcnm <- lme4::VarCorr(model_pcnm)

# Extract the species (random) effect variances and the residual variance for each model
sp_null <- as.numeric(Var_components_null[1,1])
sp_root <- as.numeric(Var_components_root[1,1])
sp_leaf <- as.numeric(Var_components_leaf[1,1])
sp_pcnm <- as.numeric(Var_components_pcnm[1,1])

res_null <- as.numeric(Var_components_null[2,1])
res_root <- as.numeric(Var_components_root[2,1])
res_leaf <- as.numeric(Var_components_leaf[2,1])
res_pcnm <- as.numeric(Var_components_pcnm[2,1])

# Obtain the total variance of the null model
v_0 <- sp_null + res_null

# Calculate the variance component linked to the root traits
var_species_root <- round(100 * sp_root / v_0)
var_res_root <- round(100 * res_root / v_0)
var_trait_root <- round(100 - (var_species_root + var_res_root)) #only linked to the traits
  
# Calculate the variance component linked to the root traits
var_species_leaf <- round(100 * sp_leaf / v_0)
var_res_leaf <- round(100 * res_leaf / v_0)
var_trait_leaf <- round(100 - (var_species_leaf + var_res_leaf)) #only linked to the traits

# Calculate the variance component linked to the location
var_species_loc <- round(100 * sp_pcnm / v_0)
var_res_loc <- round(100 * res_pcnm / v_0)
var_trait_loc <- round(100 - (var_species_loc + var_res_loc)) #only linked to the traits

#calculate the variance linked to the species identity
var_species_ID <- round(100 * sp_null / v_0)

#calculate residual variance
var_res_total <- round(100- (var_trait_leaf + var_trait_root + var_trait_loc+ var_species_ID))

#organize for plottibng
  Organ <- c(Organ, Organ, Organ, Organ, Organ)
  DiversityTrait <- c(Diversity, Diversity, Diversity, Diversity, Diversity)
  Levels <- c( "Residual", "Species", "Location", "Leaf traits", "Root traits")
  Variances <- c( var_res_total,var_species_ID, var_trait_loc, var_trait_leaf,  var_trait_root)
  
    return(data.frame(Organ, Diversity, Levels, Variances))
}

# Loop to calculate the variance partitioning in the leaves + roots
var.root.fungi <- c()

for (i in colnames(df2)[44:47]){
  
 var.root.fungi <- bind_rows( var.root.fungi, Var_par(Diversity = i, Mydata = df2 %>% filter(organ == "root"), Organ = "root"))
  
}


var.leaf.fungi <- c()

for (i in colnames(df2)[44:47]){
  
 
var.leaf.fungi <- bind_rows( var.leaf.fungi, Var_par(Diversity = i, Mydata = df2 %>% filter(organ == "leaf"), Organ = "leaf"))
  
}

vars <- rbind(var.leaf.fungi, var.root.fungi)

#plot
variance_plot<- vars %>%  
  mutate(Levels = factor(Levels, levels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Diversity)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#db7c26",  "#a7c957", "#bc4749", "#62929e", "#BBBBBB" ),
                    breaks=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"),
                    labels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)+
  ggtitle("Fungi Variance partitioning")
variance_plot
```


#data
```{r}
load("./resources/Metabarlist_natura_clean_ITS2_traits_alpha.Rdata")
```



### leaf ITS - composition
```{r}
# microbial data, hellinger transformed
leaf<- subset_metabarlist(natura_clean, table = "samples",
                          indices = natura_clean$samples$organ == "leaf")

microbio.hel <- leaf$reads %>% decostand(., "hell")
dist.compo <- vegdist(microbio.hel, method = "bray")

#trait data
names(leaf$samples)
env <- leaf$samples[,c(12,17:46 )]
# Prior to constructing the model, leaf functional trait variables were mean-centered and scaled to unit variance.
names(env)
env.scale <- scale(env[,c(2:28)], center = TRUE, scale = TRUE) #scale traits not the indices
env.scale <- as.data.frame(env.scale)
rest <- env[,c(1,29:31)] 
env.scale <- cbind(env.scale, rest)
names(env.scale)
env.scale$name <- as.factor(env.scale$name)
env.scale$name  <- unclass(env.scale$name)

#correlation between leaf traits
corr <-cor(env.scale[,-c(25:31)],method="spearman")
p.mat=cor.mtest(env.scale[,-c(28:31)],conf.level=0.95)
corrplot(corr) #nbleaflets, TLP and MAjVLA have NAs
names(env.scale)
#without nbleaflets, FW, SW, DW, RootW, ShootW, LeavesW (not pertinent), LA (since we have LA_total) and nbleaflets, TLP and MAjVLA that have NAs
env.scale <- env.scale[,-c(1,2,3,10,11,12,13,25,26,27)]
names(env.scale)
corr <-cor(env.scale[,-c(18:21)],method="spearman")
corrplot(corr)

#RDA
vare.rda <- rda(microbio.hel, env.scale, scale=T)
envfit(vare.rda, env.scale)
anova(vare.rda)

# Calculate variance partitioning
vp <- varpart(dist.compo, ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm, # leaf traits
             ~#  ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = leaf_ITS$samples)
vp
plot(vp)

root_trait_vp <- vp$part$fract[1,3] *100 
root_trait_vp <- 0
leaf_trait_vp <- vp$part$fract[2,3] *100
location_vp <- vp$part$fract[3,3] *100 #MEM 1+2+3+4
species_vp <- vp$part$fract[4,3] *100 #name
residual_vp <- 100 - (root_trait_vp +leaf_trait_vp + location_vp + species_vp) #residual

Organ <- c("Leaf")
  Phylum <- c("Fungi")
  Levels <- c("Root traits", "Leaf traits", "Location", "Species", "Residuals")
  Variances <- c(root_trait_vp,leaf_trait_vp,location_vp, species_vp, residual_vp )
vars.fungi.leaf <-   data.frame(Organ, Phylum, Levels, Variances)
vars.fungi.leaf
#plot
#vars
#plot the results
vars.fungi.leaf$Levels <- as.factor(vars.fungi.leaf$Levels)
variance_plot<- vars.fungi.leaf %>%  
  mutate(Levels = factor(Levels, levels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Phylum)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#db7c26",  "#a7c957", "#bc4749", "#62929e", "#BBBBBB" ),
                    breaks=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"),
                    labels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)+
  ggtitle("Fungi Variance partitioning")
variance_plot
```

###root ITS - matrix

```{r}
# microbial data, hellinger transformed
root_ITS <- subset_metabarlist(natura_clean, table = "samples",
                          indices = natura_clean$samples$organ == "root")

microbio.hel <- leaf_ITS$reads %>% decostand(., "hell")

# Calculate variance partitioning
vp <- varpart( decostand(root_ITS$reads, "hell"), ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm, # leaf traits
               ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = root_ITS$samples)
vp
plot(vp)

root_trait_vp <- vp$part$fract[1,3] *100 
root_trait_vp <- 0
leaf_trait_vp <- vp$part$fract[2,3] *100
location_vp <- vp$part$fract[3,3] *100 #MEM 1+2+3+4
species_vp <- vp$part$fract[4,3] *100 #name
residual_vp <- 100 - (root_trait_vp +leaf_trait_vp + location_vp + species_vp) #residual

Organ <- c("Root")
  Phylum <- c("Fungi")
  Levels <- c("Root traits", "Leaf traits", "Location", "Species", "Residuals")
  Variances <- c(root_trait_vp,leaf_trait_vp,location_vp, species_vp, residual_vp )
vars.fungi.leaf <-   data.frame(Organ, Phylum, Levels, Variances)
vars.fungi.leaf
#plot
#vars
#plot the results
vars.fungi.leaf$Levels <- as.factor(vars.fungi.leaf$Levels)
variance_plot<- vars.fungi.leaf %>%  
  mutate(Levels = factor(Levels, levels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Phylum)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#db7c26",  "#a7c957", "#bc4749", "#62929e", "#BBBBBB" ),
                    breaks=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"),
                    labels=c("Root traits", "Leaf traits", "Location", "Species", "Residuals"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)+
  ggtitle("Fungi Variance partitioning")
variance_plot
```


 Shannon for leaf
```{r}
model_null <- nlme::lme(Shannon~ 1,  random=~1|name, data = df2.leaf, na.action = na.omit, method = "ML")

model_trait <- nlme::lme(Shannon ~ RootShoot + LA_Total+ LSWC,  random=~1|name, data = df2.leaf, na.action = na.omit, method = "ML")

#don't know how to include it
model_pcnm <-  nlme::lme(microbio ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = env, na.action = na.omit, method = "ML")

 # Extract the variance components 
  Var_components <-lme4::VarCorr(model_trait)
  var_sp <- as.numeric(Var_components[1,1]) #for the random effects - species
  var_res <- as.numeric(Var_components[2,1]) #model residual
  
  # Obtain the variance of the random effect in the null model.
  random_variance_null <- as.numeric(lme4::VarCorr(model_null)[1,1])
  
  # Obtain the residual variance
  residual_variance_null <- as.numeric(lme4::VarCorr(model_null)[2,1])
  
  # Obtain the total variance of the null model
  v_0 <- random_variance_null + residual_variance_null
  
  # Calculate the variance component linked to the traits by substracting the residual variance of the model (var_res) and the variance explained by the random factor, the species (var_sp), from the variance of the null model
 
  # Variance partitioning
  var_sp <- round(100 * var_sp / v_0)
  var_res <- round(100 * var_res / v_0)
  var_trait <- round(100 - (var_sp + var_res))  # or it works also , var_envt <- v_0 - var_indv - var_sp Marion calcul
  Organ <- c("Leaf")
  Diversity <- c("Shannon")
  Levels <- c("Traits", "Species", "Residual")
  Variances <- c(var_trait,var_sp,var_res)
vars.shannon.leaf <-   data.frame(Organ, Diversity, Levels, Variances)
```

InvSimpson for leaf
```{r}
model_null <- nlme::lme(invSimpson~ 1,  random=~1|name, data = df2.leaf, na.action = na.omit, method = "ML")

model_trait <- nlme::lme(invSimpson ~ RootShoot + LA_Total+ LSWC,  random=~1|name, data = df2.leaf, na.action = na.omit, method = "ML")

#model_pcnm <-  nlme::lme(microbio ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = env, na.action = na.omit, method = "ML")

 # Extract the variance components 
  Var_components <-lme4::VarCorr(model_trait)
  var_sp <- as.numeric(Var_components[1,1]) #for the random effects - species
  var_res <- as.numeric(Var_components[2,1]) #model residual
  
  # Obtain the variance of the random effect in the null model.
  random_variance_null <- as.numeric(lme4::VarCorr(model_null)[1,1])
  
  # Obtain the residual variance
  residual_variance_null <- as.numeric(lme4::VarCorr(model_null)[2,1])
  
  # Obtain the total variance of the null model
  v_0 <- random_variance_null + residual_variance_null
  
  # Calculate the variance component linked to the traits by substracting the residual variance of the model (var_res) and the variance explained by the random factor, the species (var_sp), from the variance of the null model
 
  # Variance partitioning
  var_sp <- round(100 * var_sp / v_0)
  var_res <- round(100 * var_res / v_0)
  var_trait <- round(100 - (var_sp + var_res))  # or it works also , var_envt <- v_0 - var_indv - var_sp Marion calcul
  Organ <- c("Leaf")
  Diversity <- c("invSimpson")
  Levels <- c("Traits", "Species", "Residual")
  Variances <- c(var_trait,var_sp,var_res)
vars.invSimpson.leaf <-   data.frame(Organ, Diversity, Levels, Variances)
```

Richness for root
```{r}
df2.root<- df2 %>% filter(organ == "root")
model_null <- nlme::lme(Richness~ 1,  random=~1|name, data = df2.root, na.action = na.omit, method = "ML")

model_trait <- nlme::lme(Richness ~ RootShoot + LA_Total+ LSWC,  random=~1|name, data = df2.root, na.action = na.omit, method = "ML")

#model_pcnm <-  nlme::lme(microbio ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = env, na.action = na.omit, method = "ML")

 # Extract the variance components 
  Var_components <-lme4::VarCorr(model_trait)
  var_sp <- as.numeric(Var_components[1,1]) #for the random effects - species
  var_res <- as.numeric(Var_components[2,1]) #model residual
  
  # Obtain the variance of the random effect in the null model.
  random_variance_null <- as.numeric(lme4::VarCorr(model_null)[1,1])
  
  # Obtain the residual variance
  residual_variance_null <- as.numeric(lme4::VarCorr(model_null)[2,1])
  
  # Obtain the total variance of the null model
  v_0 <- random_variance_null + residual_variance_null
  
  # Calculate the variance component linked to the traits by substracting the residual variance of the model (var_res) and the variance explained by the random factor, the species (var_sp), from the variance of the null model
 
  # Variance partitioning
  var_sp <- round(100 * var_sp / v_0)
  var_res <- round(100 * var_res / v_0)
  var_trait <- round(100 - (var_sp + var_res))  # or it works also , var_envt <- v_0 - var_indv - var_sp Marion calcul
  Organ <- c("Root")
  Diversity <- c("Richness")
  Levels <- c("Traits", "Species", "Residual")
  Variances <- c(var_trait,var_sp,var_res)
vars.richness.root <-   data.frame(Organ, Diversity, Levels, Variances)
```

Shannon for root
```{r}
model_null <- nlme::lme(Shannon~ 1,  random=~1|name, data = df2.root, na.action = na.omit, method = "ML")

model_trait <- nlme::lme(Shannon ~ RootShoot + LA_Total+ LSWC,  random=~1|name, data = df2.root, na.action = na.omit, method = "ML")

#model_pcnm <-  nlme::lme(microbio ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = env, na.action = na.omit, method = "ML")

 # Extract the variance components 
  Var_components <-lme4::VarCorr(model_trait)
  var_sp <- as.numeric(Var_components[1,1]) #for the random effects - species
  var_res <- as.numeric(Var_components[2,1]) #model residual
  
  # Obtain the variance of the random effect in the null model.
  random_variance_null <- as.numeric(lme4::VarCorr(model_null)[1,1])
  
  # Obtain the residual variance
  residual_variance_null <- as.numeric(lme4::VarCorr(model_null)[2,1])
  
  # Obtain the total variance of the null model
  v_0 <- random_variance_null + residual_variance_null
  
  # Calculate the variance component linked to the traits by substracting the residual variance of the model (var_res) and the variance explained by the random factor, the species (var_sp), from the variance of the null model
 
  # Variance partitioning
  var_sp <- round(100 * var_sp / v_0)
  var_res <- round(100 * var_res / v_0)
  var_trait <- round(100 - (var_sp + var_res))  # or it works also , var_envt <- v_0 - var_indv - var_sp Marion calcul
  Organ <- c("Root")
  Diversity <- c("Shannon")
  Levels <- c("Traits", "Species", "Residual")
  Variances <- c(var_trait,var_sp,var_res)
vars.shannon.root <-   data.frame(Organ, Diversity, Levels, Variances)
```

InvSimpson for root
```{r}
model_null <- nlme::lme(invSimpson~ 1,  random=~1|name, data = df2.root, na.action = na.omit, method = "ML")

model_trait <- nlme::lme(invSimpson ~ RootShoot + LA_Total+ LSWC,  random=~1|name, data = df2.root, na.action = na.omit, method = "ML")

#model_pcnm <-  nlme::lme(microbio ~ MEM1 + MEM2 + MEM3 + MEM4,  random=~1|Name, data = env, na.action = na.omit, method = "ML")

 # Extract the variance components 
  Var_components <-lme4::VarCorr(model_trait)
  var_sp <- as.numeric(Var_components[1,1]) #for the random effects - species
  var_res <- as.numeric(Var_components[2,1]) #model residual
  
  # Obtain the variance of the random effect in the null model.
  random_variance_null <- as.numeric(lme4::VarCorr(model_null)[1,1])
  
  # Obtain the residual variance
  residual_variance_null <- as.numeric(lme4::VarCorr(model_null)[2,1])
  
  # Obtain the total variance of the null model
  v_0 <- random_variance_null + residual_variance_null
  
  # Calculate the variance component linked to the traits by substracting the residual variance of the model (var_res) and the variance explained by the random factor, the species (var_sp), from the variance of the null model
 
  # Variance partitioning
  var_sp <- round(100 * var_sp / v_0)
  var_res <- round(100 * var_res / v_0)
  var_trait <- round(100 - (var_sp + var_res))  # or it works also , var_envt <- v_0 - var_indv - var_sp Marion calcul
  Organ <- c("Root")
  Diversity <- c("invSimpson")
  Levels <- c("Traits", "Species", "Residual")
  Variances <- c(var_trait,var_sp,var_res)
vars.invSimpson.root <-   data.frame(Organ, Diversity, Levels, Variances)
```



Plot
```{r}
vars <- rbind(vars.richness.leaf, vars.shannon.leaf, vars.invSimpson.leaf, vars.richness.root, vars.shannon.root, vars.invSimpson.root)
#vars
#plot the results
vars$Levels <- as.factor(vars$Levels)
variance_plot<- vars %>%  
  mutate(Levels = factor(Levels, levels=c("Traits", "Species", "Residual"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Diversity)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#029A88",  "#CCBC44", "#BBBBBB" ),
                    breaks=c("Traits", "Species", "Residual" ),
                    labels=c("Traits", "Species", "Residual" ))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)+
  ggtitle("16S Variance partitioning")
variance_plot
```

Relative contribution of the different predictors used to model bacterial and fungal composition and diversity. Panels represent results from variation partitioning modelling aiming to identify the percentage variance of bacterial and fungal community composition and diversity explained by plant attributes (RootShoot + LA_Total+ LSWC), species identity and climate in Australia and England. Unique and shared variance from plant cover, diversity and functional traits in predicting microbial community composition and diversity were merged in this figure for simplicity.


