---
title: "Variance partitioning"
author: "Marion Boisseaux"
date: "2023-05-10"
output: html_document
---

**Variance partitioning** 

# library
```{r}
library(vegan)
```


(rda(spe_hellinger, env_water_chemistry, env_topograhpy))
Test fraction [c]
anova.cca(rda(spe_hellinger, env_topograhpy, env_water_chemistry))

# leaf ITS 

```{r}
load("./resources/Metabarlist_natura_clean_ITS2_traits_alpha.Rdata")
leaf <- subset_metabarlist(natura_clean, table = "samples",
                          indices = natura_clean$samples$organ == "leaf")
```


## vp on Shannon

```{r}

#trait data
trait_st <- leaf$samples %>% dplyr::select(SLA, LA_rwc, RootShoot, RootLength_Total, Height, LSWC, StomatalDensity, SRL, RootDiameter, LT, gmin) %>% scale(center = TRUE, scale = TRUE) %>% as.data.frame()# Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.
trait_st$sample_id <- rownames(trait_st)
trait_st <- left_join(trait_st, leaf$samples %>% dplyr::select(sample_id, name))
names(trait_st)
Shannon <- leaf$samples %>% dplyr::select(Shannon, sample_id)
trait_st <- left_join(trait_st, Shannon)
names(trait_st)


# Calculate variance partitioning
vp <-varpart(Y = leaf$samples$Shannon ,  ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + StomatalDensity + SLA + LA_rwc, # leaf traits
             ~Height, #plant trait
              # ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = trait_st)
vp
showvarparts(4, bg=2:4)
plot(vp, digits = 3, Xnames = c("root traits", "leaf traits","plant traits", "species Id"))

# 
# #use RDA to test significance of fractions of interest
# 
# # Test fraction [a] using partial RDA:
# roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, trait_st)[,-1]
# # Test fraction [b] using partial RDA:
# leaftraits <- model.matrix(~LSWC + LT + gmin + StomatalDensity + fvfm, trait_st)[,-1]
# # Test fraction [c] using partial RDA:
# location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, trait_st)[,-1]
# # Test fraction [d] using partial RDA:
# identity <- model.matrix(~name, trait_st)[,-1]
# # Test fraction [e] using partial RDA:
# planttraits <- model.matrix(~Height + NbLeaves + LA_Total, trait_st)[,-1]
# 
# aFrac <- rda(trait_st$Shannon, roottraits)
# bFrac <- rda(trait_st$Shannon, leaftraits)
# cFrac <- rda(trait_st$Shannon, location)
# dFrac <- rda(trait_st$Shannon, identity)
# eFrac <- rda(trait_st$Shannon, planttraits)
# 
# 
# #Test fraction with anovo or anova.cca
# anova(aFrac, step=200, perm.max=200) #0.007 ** 
# anova(bFrac, step=200, perm.max=200) #0.139
# anova(cFrac, step=200, perm.max=200) #0.863
# anova(dFrac, step=200, perm.max=200) #0.001 ***
# anova(eFrac, step=200, perm.max=200) #0.001 ***
# 
# # RsquareAdj gives the same result as component [a] of varpart
# a <- unlist(RsquareAdj(aFrac)[2]) #15% X1
# b <- unlist(RsquareAdj(bFrac)[2]) #34% X2
# c <- unlist(RsquareAdj(cFrac)[2]) #0% X3
# d <- unlist(RsquareAdj(dFrac)[2]) #40% X4
# 
# total <- a+b+d
# 
# total
# vp <-varpart(Y = leaf$samples$Shannon , 
#              ~ NbLeaves + LA_Total + Height, #plant traits
#                ~name, #species' ID
#                 ~ SRL + RootDiameter + RootLength_Total, #location
#                data = trait_st)
# plot(vp)
```

Plot les effets **purs**
```{r}
 # prepare plot for diversity 
  vp_leaf <- 0
  vp_root <- round(vp$part$indfract[3,3] *100)
  vp_plant <- round(vp$part$indfract[1,3] *100)
  vp_name <- round(vp$part$indfract[2,3] *100)
  vp_location <- 0
  vp_shared <- round(sum(vp$part$indfract[4:7,3]) *100)
  vp_res <- 100- (vp_root + vp_name + vp_plant + vp_shared)

  Organ <- c("Leaf")
  Microorganism <- c("Fungi")
  Diversity <- c("Shannon")
  Levels <- c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual")
  Variances <- c(vp_leaf, vp_root, vp_plant, vp_name, vp_location, vp_shared, vp_res)
  data.plot <-  data.frame(Organ, Microorganism, Diversity, Levels, Variances)
  
  #plot the results
 data.plot$Levels <- as.factor(data.plot$Levels)
 variance_plot<- data.plot %>% 
  mutate(Levels = factor(Levels, levels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Microorganism)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#a7c957","#d4a373", "#a8dadc",  "#ffb4a2","#ffba08", "#BBBBBB", "#f8f9fa" ),
                    breaks=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"),
                    labels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)
variance_plot -> vp_leaf_fungi_diversity
vp_leaf_fungi_diversity
```


## vp on composition

Bray–Curtis distances were used for microbial composition in this analysis.

```{r}
# microbial data, hellinger transformed or not
microbio.hel <- leaf$reads %>% decostand(., "hell")
microbio.hel.bray <- vegdist(microbio.hel, method = "bray")
microbio <- vegdist(leaf$reads, method = "bray")

#trait data
trait_st <- leaf$samples %>% dplyr::select(SLA, LA_rwc, RootShoot, RootLength_Total, Height, LSWC, StomatalDensity, SRL, RootDiameter, LT, gmin) %>% scale(center = TRUE, scale = TRUE) %>% as.data.frame()# Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.
trait_st$sample_id <- rownames(trait_st)
trait_st <- left_join(trait_st,leaf$samples %>% dplyr::select(sample_id, name))
names(trait_st)
trait_st$name  <- as.factor(trait_st$name)
trait_st$name  <- unclass(trait_st$name)

# Calculate variance partitioning
vp2 <-varpart(Y = microbio.hel.bray ,  
              # ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + StomatalDensity + SLA + LA_rwc, # leaf traits
              # ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = trait_st)
vp2
showvarparts(2, bg=2:4)
plot(vp2, digits = 3, Xnames = c( "leaf traits","species Id"))


#Use function ‘dbrda’ to test significance of fractions of interest

# Test fraction [a] using partial RDA:
roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, trait_st)[,-1]
# Test fraction [b] using partial RDA:
leaftraits <- model.matrix(~LSWC + LT + gmin + StomatalDensity + fvfm, trait_st)[,-1]
# Test fraction [c] using partial RDA:
location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, trait_st)[,-1]
# Test fraction [d] using partial RDA:
identity <- model.matrix(~name, trait_st)[,-1]
# Test fraction [e] using partial RDA:
planttraits <- model.matrix(~Height + NbLeaves + LA_Total, trait_st)[,-1]

# aFrac <- dbrda(microbio.hel.bray ~ roottraits)
# bFrac <- dbrda(microbio.hel.bray ~  leaftraits)
# cFrac <- dbrda(microbio.hel.bray ~  location)
# dFrac <- dbrda(microbio.hel.bray~ identity)
# eFrac <- dbrda(microbio.hel.bray~ planttraits)

aFrac <- dbrda(microbio ~ roottraits)
bFrac <- dbrda(microbio ~  leaftraits)
cFrac <- dbrda(microbio ~  location)
dFrac <- dbrda(microbio~ identity)
eFrac <- dbrda(microbio~ planttraits)
all <- capscale(microbio~ roottraits + leaftraits + location + identity)
all2 <- dbrda(leaf$reads~ roottraits + leaftraits, distance = "bray")
anova(all2, step=200, perm.max=200, by= "terms" ) #*
plot(all)
global_r2 <- RsquareAdj(all2)$adj.r.squared # 4%
sqrt(vif.cca(all2)) #As a rule of the thumb, if VIF−−−−√>2 multicollinearity is considered high

plot(all2, type= "n", scaling= "species")
points(all2, display="sites", pch=20)
text(all2$terms, display="bp", col=4)
text(bFrac, display="bp", col=3 )

null_model <- dbrda(microbio ~ 1 , data = trait_st) 

a <- ordiR2step (null_model, scope = formula(all2), direction = c("forward"))
a
(microbio, # Y matrix
  trait_st, # X matrix
  adjR2thresh = global_r2, # Set the adj.R2 threshold
  alpha = 0.001, # Set alpha level
  nperm = 999  # Number of permutations
)   

Stop generating

0.04487*0.28 = 0.0125

anova(aFrac, step=200, perm.max=200) #0.695
anova(bFrac, step=200, perm.max=200) # 0.011 *
anova(cFrac, step=200, perm.max=200) #  0.011 *
anova(dFrac, step=200, perm.max=200) # 0.008 **
anova(eFrac, step=200, perm.max=200) #   0.477

# RsquareAdj gives the same result as component [a] of varpart
a <- unlist(RsquareAdj(aFrac)[2])
b <- unlist(RsquareAdj(bFrac)[2])
c <- unlist(RsquareAdj(cFrac)[2])
d <- unlist(RsquareAdj(dFrac)[2])

#final model
vp2 <-varpart(Y = microbio ,  
               ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + StomatalDensity + fvfm, # leaf traits
               #~ MEM1 + MEM2 + MEM3 +MEM4, #location
               #~name, #species' ID
              #~Height + NbLeaves + LA_Total,  
               data = trait_st)
vp2
plot(vp2)
```

[a+d+f+g] = X1         5  0.08320      0.01268     TRUE
[b+d+e+g] = X2         4  0.07129      0.01501     TRUE
[c+e+f+g] = X3         6  0.10211      0.01794     TRUE

Plot les effets **purs**
```{r}
 # prepare plot for diversity 
  vp_leaf <- round( 0.01268  *100)
  vp_root <- 0
  vp_plant <- 0
  vp_name <- round( 0.01794  *100)
  vp_location <- round(0.01501 *100)
  vp_shared <- round(sum(vp2$part$indfract[4:7,3]) *100)
  vp_res <- 100- (vp_leaf + vp_plant + vp_name + vp_location + vp_shared)

  Organ <- c("Leaf")
  Microorganism <- c("Fungi")
  Diversity <- c("Composition")
  Levels <- c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared")
  Variances <- c(vp_leaf, vp_root, vp_plant, vp_name, vp_location, vp_shared)
  data.plot.2 <-  data.frame(Organ, Microorganism, Diversity, Levels, Variances)
  
  #plot the results
 data.plot2$Levels <- as.factor(data.plot$Levels)
 variance_plot<- data.plot.2 %>% 
  mutate(Levels = factor(Levels, levels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Microorganism)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#a7c957","#d4a373", "#a8dadc",  "#ffb4a2","#ffba08", "#BBBBBB"),
                    breaks=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared"),
                    labels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)
variance_plot -> vp_leaf_fungi_compo
vp_leaf_fungi_compo
```




#merge for leaf ITS compo + dv

```{r}
leaf_ITS <- rbind(data.plot, data.plot.2)
#plot

 leaf_ITS %>% 
  mutate(Levels = factor(Levels, levels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Microorganism)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#a7c957","#d4a373", "#a8dadc",  "#ffb4a2","#ffba08", "#BBBBBB", "#f8f9fa" ),
                    breaks=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"),
                    labels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Diversity)
```

# root ITS 

```{r}
load("./resources/Metabarlist_natura_clean_ITS2_traits_alpha.Rdata")
root <- subset_metabarlist(natura_clean, table = "samples",
                          indices = natura_clean$samples$organ == "root")
```


## vp on Shannon

```{r}

#trait data
trait_st <- root$samples %>% dplyr::select(RootShoot, RootLength_Total, LA_Total, Height, StemDiameter, NbLeaves, LSWC, fvfm, StomatalDensity, SRL, RootDiameter, LT, gmin, MEM1, MEM2, MEM3, MEM4, MajVLA, TLP) %>% scale(center = TRUE, scale = TRUE) %>% as.data.frame()# Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.
trait_st$sample_id <- rownames(trait_st)
trait_st <- left_join(trait_st, root$samples %>% dplyr::select(sample_id, name))
names(trait_st)
Shannon <- root$samples %>% dplyr::select(Shannon, sample_id)
trait_st <- left_join(trait_st, Shannon)
names(trait_st)


# Calculate variance partitioning
vp3 <-varpart(Y = root$samples$Shannon ,  ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm, # leaf traits
               ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = trait_st)
vp3
showvarparts(3, bg=2:4)
plot(vp, digits = 3, Xnames = c("root traits", "leaf traits","species Id"))


#use RDA to test significance of fractions of interest

# Test fraction [a] using partial RDA:
roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, trait_st)[,-1]
# Test fraction [b] using partial RDA:
leaftraits <- model.matrix(~LSWC + LT + gmin + StomatalDensity + fvfm, trait_st)[,-1]
# Test fraction [c] using partial RDA:
location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, trait_st)[,-1]
# Test fraction [d] using partial RDA:
identity <- model.matrix(~name, trait_st)[,-1]
# Test fraction [e] using partial RDA:
planttraits <- model.matrix(~Height + NbLeaves + LA_Total, trait_st)[,-1]

aFrac <- rda(trait_st$Shannon, roottraits)
bFrac <- rda(trait_st$Shannon, leaftraits)
cFrac <- rda(trait_st$Shannon, location)
dFrac <- rda(trait_st$Shannon, identity)
eFrac <- rda(trait_st$Shannon, planttraits)


#Test fraction with anovo or anova.cca
anova(aFrac, step=200, perm.max=200) # 0.76 root
anova(bFrac, step=200, perm.max=200) # 0.067 . leaf
anova(cFrac, step=200, perm.max=200) # 0.592 location
anova(dFrac, step=200, perm.max=200) # 0.002 ** identity
anova(eFrac, step=200, perm.max=200) #0.101 planttraits

vp3 <-varpart(Y = root$samples$Shannon , 
             ~ NbLeaves + LA_Total + Height, #plant traits
               ~name, #species' ID
                ~ LSWC + LT + gmin + StomatalDensity + fvfm,
               data = trait_st)
vp3
```
Individual fractions                                   
[a] = X1 | X2+X3       3               0.00238     TRUE plant
[b] = X2 | X1+X3       6               0.03052     TRUE name
[c] = X3 | X1+X2       5              -0.04631     TRUE leaf

Plot les effets **purs**
```{r}
 # prepare plot for diversity 
  vp_leaf <- 0
  vp_root <- 0
  vp_plant <- round(vp3$part$indfract[1,3] *100) #0
  vp_name <- round(vp$part$indfract[2,3] *100)
  vp_location <- 0
  vp_shared <- round(sum(vp$part$indfract[4:7,3]) *100)
  vp_res <- 100- (vp_root + vp_leaf + vp_name + vp_plant + vp_location + vp_shared)

  Organ <- c("Root")
  Microorganism <- c("Fungi")
  Diversity <- c("Shannon")
  Levels <- c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual")
  Variances <- c(vp_leaf, vp_root, vp_plant, vp_name, vp_location, vp_shared, vp_res)
  data.plot.3 <-  data.frame(Organ, Microorganism, Diversity, Levels, Variances)
  
  #plot the results
 data.plot$Levels <- as.factor(data.plot$Levels)
 variance_plot<- data.plot.3 %>% 
  mutate(Levels = factor(Levels, levels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Microorganism)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#a7c957","#d4a373", "#a8dadc",  "#ffb4a2","#ffba08", "#BBBBBB", "#f8f9fa" ),
                    breaks=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"),
                    labels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Organ)
```

## vp on composition

Bray–Curtis distances were used for microbial composition in this analysis.

```{r}
# microbial data, hellinger transformed
microbio.hel <- root$reads %>% decostand(., "hell")
microbio.hel.bray <- vegdist(microbio.hel, method = "bray")

#trait data
trait_st <- root$samples %>% dplyr::select(RootShoot, RootLength_Total, LA_Total, Height, StemDiameter, NbLeaves, LSWC, fvfm, StomatalDensity, SRL, RootDiameter, LT, gmin, MEM1, MEM2, MEM3, MEM4, MajVLA, TLP) %>% scale(center = TRUE, scale = TRUE) %>% as.data.frame()# Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.
trait_st$sample_id <- rownames(trait_st)
trait_st <- left_join(trait_st, root$samples %>% dplyr::select(sample_id, name))
names(trait_st)

# Calculate variance partitioning
vp4 <-varpart(Y = microbio.hel.bray ,  
               ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin +  StomatalDensity + fvfm, # leaf traits
               ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = trait_st)
vp4
showvarparts(3, bg=2:4)

#Use function ‘dbrda’ to test significance of fractions of interest

# Test fraction [a] using partial RDA:
roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, trait_st)[,-1]
# Test fraction [b] using partial RDA:
leaftraits <- model.matrix(~LSWC + LT + gmin + StomatalDensity + fvfm, trait_st)[,-1]
# Test fraction [c] using partial RDA:
location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, trait_st)[,-1]
# Test fraction [d] using partial RDA:
identity <- model.matrix(~name, trait_st)[,-1]
# Test fraction [e] using partial RDA:
planttraits <- model.matrix(~Height + NbLeaves + LA_Total, trait_st)[,-1]

aFrac <- dbrda(microbio.hel.bray ~ roottraits)
bFrac <- dbrda(microbio.hel.bray ~  leaftraits)
cFrac <- dbrda(microbio.hel.bray ~  location)
dFrac <- dbrda(microbio.hel.bray~ identity)
eFrac <- dbrda(microbio.hel.bray~ planttraits)

anova(aFrac, step=200, perm.max=200) #0.675
anova(bFrac, step=200, perm.max=200) # 0.493
anova(cFrac, step=200, perm.max=200) # 0.337
anova(dFrac, step=200, perm.max=200) #  0.105
anova(eFrac, step=200, perm.max=200) # 0.104

# RsquareAdj gives the same result as component [a] of varpart
a <- unlist(RsquareAdj(aFrac)[2])
b <- unlist(RsquareAdj(bFrac)[2])
c <- unlist(RsquareAdj(cFrac)[2])
d <- unlist(RsquareAdj(dFrac)[2])


```


#merge for leaf ITS compo + dv

```{r}
leaf_ITS.2 <- rbind(leaf_ITS, data.plot.3, data.plot.4)
#plot

ITS.vp <- leaf_ITS.2 %>% 
  mutate(Levels = factor(Levels, levels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))) %>%
  ggplot(aes(fill=Levels, y=Variances, x=Organ)) + 
  geom_bar(position="stack", stat="identity") +
  theme_minimal(base_size = 22) +
  ylab("") + xlab("")+
  theme(legend.text = element_text(face = "italic"),legend.position = "bottom") +
  scale_fill_manual("", values=c("#a7c957","#d4a373", "#a8dadc",  "#ffb4a2","#ffba08", "#BBBBBB", "#f8f9fa" ),
                    breaks=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"),
                    labels=c("Leaf traits","Root traits","Plant traits", "Species","Location", "Shared", "Residual"))  +
  coord_flip() +
  scale_x_discrete(labels = scales::label_parse())+
  facet_wrap(~Diversity)+
     ggtitle("Fungi - Variance partitioning")

ggsave("./results/var_part_ITS.jpeg", ITS.vp)
```

# XXXXXXX

# leaf 16S 

```{r}
load("./resources/Metabarlist_natura_clean_16S_traits_alpha.Rdata")
leaf <- subset_metabarlist(natura_clean, table = "samples",
                          indices = natura_clean$samples$organ == "leaf")
```


## vp on Shannon

```{r}

#trait data
trait_st <- leaf$samples %>% dplyr::select(RootShoot, RootLength_Total, LA_Total, Height, StemDiameter, NbLeaves, LSWC, fvfm, StomatalDensity, SRL, RootDiameter, LT, gmin, MEM1, MEM2, MEM3, MEM4, MajVLA, TLP) %>% scale(center = TRUE, scale = TRUE) %>% as.data.frame()# Each value in a column is standardized to a mean of 0 and standard deviation of 1. That is, each of the variables are normalized. This is another way to adjust for differences in the spread of values.
trait_st$sample_id <- rownames(trait_st)
trait_st <- left_join(trait_st, leaf$samples %>% dplyr::select(sample_id, name))
names(trait_st)
Shannon <- leaf$samples %>% dplyr::select(Shannon, sample_id)
trait_st <- left_join(trait_st, Shannon)
names(trait_st)


# Calculate variance partitioning
vp <-varpart(Y = leaf$samples$Shannon ,  ~SRL + RootDiameter + RootLength_Total , #root traits
               ~LSWC + LT + gmin + NbLeaves + LA_Total + StomatalDensity + fvfm, # leaf traits
               ~ MEM1 + MEM2 + MEM3 +MEM4, #location
               ~name, #species' ID
               data = trait_st)
vp
showvarparts(3, bg=2:4)
plot(vp, digits = 3, Xnames = c("root traits", "leaf traits","species Id"))


#use RDA to test significance of fractions of interest

# Test fraction [a] using partial RDA:
roottraits <- model.matrix(~SRL + RootDiameter + RootLength_Total, trait_st)[,-1]
# Test fraction [b] using partial RDA:
leaftraits <- model.matrix(~LSWC + LT + gmin + StomatalDensity + fvfm, trait_st)[,-1]
# Test fraction [c] using partial RDA:
location <- model.matrix(~MEM1 + MEM2 + MEM3 +MEM4, trait_st)[,-1]
# Test fraction [d] using partial RDA:
identity <- model.matrix(~name, trait_st)[,-1]
# Test fraction [e] using partial RDA:
planttraits <- model.matrix(~Height + NbLeaves + LA_Total, trait_st)[,-1]

aFrac <- rda(trait_st$Shannon, roottraits)
bFrac <- rda(trait_st$Shannon, leaftraits)
cFrac <- rda(trait_st$Shannon, location)
dFrac <- rda(trait_st$Shannon, identity)
eFrac <- rda(trait_st$Shannon, planttraits)


#Test fraction with anovo or anova.cca
anova(aFrac, step=200, perm.max=200) #0.058 .
anova(bFrac, step=200, perm.max=200) # 0.001 ***
anova(cFrac, step=200, perm.max=200) #0.264
anova(dFrac, step=200, perm.max=200) #0.001 ***
anova(eFrac, step=200, perm.max=200) #0.01 **

vp <-varpart(Y = leaf$samples$Shannon , 
             ~LSWC + LT + gmin + StomatalDensity + fvfm,
            
               ~name, #species' ID
                 ~ NbLeaves + LA_Total + Height, #plant traits
               data = trait_st)
vp
```

Individual fractions                                   
[a] = X1 | X2+X3       5               0.00027     TRUE leaf
[b] = X2 | X1+X3       6               0.00852     TRUE name
[c] = X3 | X1+X2       3              -0.01503     TRUE plant traits
[